`SCPCL001114` - Exploration of gene set based methods for tumor cell assignment

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_singler-classification.Rmd
Styled code

# compare blueprint to tumor, original and original + blueprint
plot_jaccard(classification_df,
             annotation_column = "blueprint_tumor_classification_lumped",
             methods_to_compare = methods,
             column_title = "Blueprint",
             legend_match = "Original") |> 
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )

@@

plot_jaccard(classification_df,
             annotation_column = "updated_singler_classification_lumped",
             methods_to_compare = subset_methods,
             column_title = "Original",
             legend_match = "Original")

# compare tumor to original and blueprint
plot_jaccard(classification_df,
             annotation_column = "tumor_only_classification_lumped",
             methods_to_compare = c("Original and Blueprint" = "updated_singler_blueprint_classification_lumped"),
             column_title = "Tumor only",
             legend_match = "Tumor only")

```

@@

```{r}
singler_annotation_columns |> 
  purrr::map(\(classification_column){

    # first get list of tumor cells for this classificationn
    # we want to remove these since we are only interested in comparing normal cell types
    tumor_cells <- classification_df |> 
      dplyr::filter({{classification_column}} == "Tumor") |> 
      dplyr::pull(barcodes)

@@

    caret_df <- classification_df |> 
      # first remove tumor cells 
      dplyr::filter(!barcodes %in% tumor_cells) |> 
      # then update to make sure all cell types are shared
      # anything not in shared cell types are classified as "other" 
      dplyr::mutate(singler_caret = dplyr::if_else(singler_celltype_annotation %in% shared_celltypes, singler_celltype_annotation, "Other"),
                    new_ref_caret = dplyr::if_else(!!sym(classification_column) %in% shared_celltypes, !!sym(classification_column), "Other"))

    caret::confusionMatrix(
      table(
        caret_df$singler_caret, 
        caret_df$new_ref_caret)
    )

  })
```

@@

```{r}
# make a table showing which cells are assigned to which reference 
celltype_table <- table(classification_df$updated_singler_blueprint_classification, 
      classification_df$orig_ref)
# sort by SCPCL000822
celltype_table[order(celltype_table[,2], decreasing = TRUE), ]
```

@@

# heatmap comparing each singler annotation type to each of the marker gene based methods 
heatmap_list <- singler_annotation_columns |> 
  purrr::iwalk(\(annotation_column, name) {

    # first make singler annotations normal or tumor for easier visualization
    plot_df <- classification_df |> 
      dplyr::mutate(new_annotation = dplyr::if_else(!!sym(annotation_column) == "Tumor", "Tumor", "Normal"))

@@

ggplot(marker_density_df, aes(x = sum_transformed_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method), 
             scales = "free_y") +
  labs(
    title = "Marker gene expression",
    x = "Sum of marker genes"
  )
```

@@

ggplot(geneset_plot_df, aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset),
             scales = "free_y") +
  labs(
    title = "Gene set scores",
    x = "Mean gene set score"
  )

```

>>>>

# compare blueprint to tumor, original and original + blueprint
plot_jaccard(classification_df,
  annotation_column = "blueprint_tumor_classification_lumped",
  methods_to_compare = methods,
  column_title = "Blueprint",
  legend_match = "Original"
) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )

@@

plot_jaccard(classification_df,
  annotation_column = "updated_singler_classification_lumped",
  methods_to_compare = subset_methods,
  column_title = "Original",
  legend_match = "Original"
)

# compare tumor to original and blueprint
plot_jaccard(classification_df,
  annotation_column = "tumor_only_classification_lumped",
  methods_to_compare = c("Original and Blueprint" = "updated_singler_blueprint_classification_lumped"),
  column_title = "Tumor only",
  legend_match = "Tumor only"
)
```

@@

```{r}
singler_annotation_columns |>
  purrr::map(\(classification_column){
    # first get list of tumor cells for this classificationn
    # we want to remove these since we are only interested in comparing normal cell types
    tumor_cells <- classification_df |>
      dplyr::filter({{ classification_column }} == "Tumor") |>
      dplyr::pull(barcodes)

@@

    caret_df <- classification_df |>
      # first remove tumor cells
      dplyr::filter(!barcodes %in% tumor_cells) |>
      # then update to make sure all cell types are shared
      # anything not in shared cell types are classified as "other"
      dplyr::mutate(
        singler_caret = dplyr::if_else(singler_celltype_annotation %in% shared_celltypes, singler_celltype_annotation, "Other"),
        new_ref_caret = dplyr::if_else(!!sym(classification_column) %in% shared_celltypes, !!sym(classification_column), "Other")
      )

    caret::confusionMatrix(
      table(
        caret_df$singler_caret,
        caret_df$new_ref_caret
      )
    )
  })
```

@@

```{r}
# make a table showing which cells are assigned to which reference
celltype_table <- table(
  classification_df$updated_singler_blueprint_classification,
  classification_df$orig_ref
)
# sort by SCPCL000822
celltype_table[order(celltype_table[, 2], decreasing = TRUE), ]
```

@@

# heatmap comparing each singler annotation type to each of the marker gene based methods
heatmap_list <- singler_annotation_columns |>
  purrr::iwalk(\(annotation_column, name) {
    # first make singler annotations normal or tumor for easier visualization
    plot_df <- classification_df |>
      dplyr::mutate(new_annotation = dplyr::if_else(!!sym(annotation_column) == "Tumor", "Tumor", "Normal"))

@@

ggplot(marker_density_df, aes(x = sum_transformed_exp, color = classification)) +
  geom_density() +
  facet_wrap(vars(method),
    scales = "free_y"
  ) +
  labs(
    title = "Marker gene expression",
    x = "Sum of marker genes"
  )
```

@@

ggplot(geneset_plot_df, aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(
    rows = vars(method),
    cols = vars(geneset),
    scales = "free_y"
  ) +
  labs(
    title = "Gene set scores",
    x = "Mean gene set score"
  )
```

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000752/SCPCL001114_tumor-gene-expression-exploration.Rmd
Styled code

ref_markers_df <- create_marker_gene_df(sce = ref_sce, 
                                        classification_df = ref_classification_df,
                                        marker_genes_file) |> 
  dplyr::mutate(sample = "SCPCL000822")

@@

```{r}
# create distribution 
density_data <- density(combined_markers_df$sum_raw_exp)
# find the local minima in the distribution
exp_cutoff <- optimize(approxfun(density_data$x, density_data$y), interval = c(1,10))$minimum
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$ref_marker_gene_classification, 
      classification_df$auc_classification)
  ) 
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$ref_marker_gene_classification,
      classification_df$one_marker_gene_classification)
  ) 
```

@@

classification_df <- classification_df |> 
  dplyr::mutate(
    two_marker_gene_classification = dplyr::if_else(
      barcodes %in% tumor_cells,
      "Tumor",
      "Normal"
    ))
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$ref_marker_gene_classification,
      classification_df$two_marker_gene_classification)
  ) 
```

@@

ucell_df <- new_ucell |> 
  as.data.frame() |> 
  tibble::rownames_to_column("barcodes") |> 
  dplyr::mutate(sample = "SCPCL001114") |> 
  dplyr::bind_rows(ref_ucell) |> 
  tidyr::pivot_longer(ends_with("UCell"), 
                      names_to = "gene_list", 
                      values_to = "signature_score") |> 
  dplyr::mutate(
    gene_list = stringr::word(gene_list, 1, sep = "_")
  )
```

```{r}
ggplot(ucell_df, aes(x = signature_score, colour = sample)) +
  geom_density() +
  facet_grid(rows = vars(gene_list),
             scales = "free_y")
```

@@

classification_df |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = forcats::fct_recode(method, !!!method_levels),
    method = forcats::fct_relevel(method, "Consensus", after = Inf)) |> 
  ggplot(aes(x = UMAP1, y = UMAP2, color = classification)) +
  geom_point(size = 0.1, alpha = 0.5) +
  facet_wrap(vars(method)) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
```

@@

```{r fig.height=7}
geneset_plot_df |> 
  # add new methods 
  dplyr::left_join(new_classification_df, by = c("barcodes", "ref_marker_gene_classification")) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = forcats::fct_recode(method, !!!method_levels), 
    method = forcats::fct_relevel(method, "Consensus", after = Inf)) |> 
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))
```

@@

annotation <- ComplexHeatmap::columnAnnotation(
  one_marker = annotation_df$one_marker_gene_classification,
  two_marker = annotation_df$two_marker_gene_classification,
  ref_marker_genes = annotation_df$ref_marker_gene_classification,
  AUCell = annotation_df$auc_classification, 
  consensus = annotation_df$consensus_classification,
  col = list(one_marker = plot_colors,
             two_marker = plot_colors,
             ref_marker_genes = plot_colors, 
             AUCell = plot_colors,
             consensus = plot_colors))
```

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- plot_markers_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

# plot heatmap of marker genes 
plot_gene_heatmap(marker_gene_heatmap, 
                  row_title = "Marker gene symbol",
                  legend_title = "Marker gene \nexpression",
                  annotation = annotation)
```

```{r}
# make a matrix of gene set by barcode 
geneset_heatmap <- geneset_plot_df |> 
  dplyr::select(mean_score, geneset, barcodes) |> 
  unique() |> 
  tidyr::pivot_wider(values_from = mean_score, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("geneset") |> 
  as.matrix()

# plot heatmap of gene set score 
plot_gene_heatmap(geneset_heatmap,
                  annotation = annotation,
                  legend_title = "Gene set \nscore")
```

>>>>

ref_markers_df <- create_marker_gene_df(
  sce = ref_sce,
  classification_df = ref_classification_df,
  marker_genes_file
) |>
  dplyr::mutate(sample = "SCPCL000822")

@@

```{r}
# create distribution
density_data <- density(combined_markers_df$sum_raw_exp)
# find the local minima in the distribution
exp_cutoff <- optimize(approxfun(density_data$x, density_data$y), interval = c(1, 10))$minimum
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
  table(
    classification_df$ref_marker_gene_classification,
    classification_df$auc_classification
  )
)
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
  table(
    classification_df$ref_marker_gene_classification,
    classification_df$one_marker_gene_classification
  )
)
```

@@

classification_df <- classification_df |>
  dplyr::mutate(
    two_marker_gene_classification = dplyr::if_else(
      barcodes %in% tumor_cells,
      "Tumor",
      "Normal"
    )
  )
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
  table(
    classification_df$ref_marker_gene_classification,
    classification_df$two_marker_gene_classification
  )
)
```

@@

ucell_df <- new_ucell |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::mutate(sample = "SCPCL001114") |>
  dplyr::bind_rows(ref_ucell) |>
  tidyr::pivot_longer(ends_with("UCell"),
    names_to = "gene_list",
    values_to = "signature_score"
  ) |>
  dplyr::mutate(
    gene_list = stringr::word(gene_list, 1, sep = "_")
  )
```

```{r}
ggplot(ucell_df, aes(x = signature_score, colour = sample)) +
  geom_density() +
  facet_grid(
    rows = vars(gene_list),
    scales = "free_y"
  )
```

@@

classification_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |>
  dplyr::mutate(
    method = forcats::fct_recode(method, !!!method_levels),
    method = forcats::fct_relevel(method, "Consensus", after = Inf)
  ) |>
  ggplot(aes(x = UMAP1, y = UMAP2, color = classification)) +
  geom_point(size = 0.1, alpha = 0.5) +
  facet_wrap(vars(method)) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
```

@@

```{r fig.height=7}
geneset_plot_df |>
  # add new methods
  dplyr::left_join(new_classification_df, by = c("barcodes", "ref_marker_gene_classification")) |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |>
  dplyr::mutate(
    method = forcats::fct_recode(method, !!!method_levels),
    method = forcats::fct_relevel(method, "Consensus", after = Inf)
  ) |>
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(
    rows = vars(method),
    cols = vars(geneset)
  )
```

@@

annotation <- ComplexHeatmap::columnAnnotation(
  one_marker = annotation_df$one_marker_gene_classification,
  two_marker = annotation_df$two_marker_gene_classification,
  ref_marker_genes = annotation_df$ref_marker_gene_classification,
  AUCell = annotation_df$auc_classification,
  consensus = annotation_df$consensus_classification,
  col = list(
    one_marker = plot_colors,
    two_marker = plot_colors,
    ref_marker_genes = plot_colors,
    AUCell = plot_colors,
    consensus = plot_colors
  )
)
```

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- plot_markers_df |>
  dplyr::select(gene_expression, gene_symbol, barcodes) |>
  tidyr::pivot_wider(
    values_from = gene_expression,
    names_from = barcodes
  ) |>
  tibble::column_to_rownames("gene_symbol") |>
  as.matrix()

# plot heatmap of marker genes
plot_gene_heatmap(marker_gene_heatmap,
  row_title = "Marker gene symbol",
  legend_title = "Marker gene \nexpression",
  annotation = annotation
)
```

```{r}
# make a matrix of gene set by barcode
geneset_heatmap <- geneset_plot_df |>
  dplyr::select(mean_score, geneset, barcodes) |>
  unique() |>
  tidyr::pivot_wider(
    values_from = mean_score,
    names_from = barcodes
  ) |>
  tibble::column_to_rownames("geneset") |>
  as.matrix()

# plot heatmap of gene set score
plot_gene_heatmap(geneset_heatmap,
  annotation = annotation,
  legend_title = "Gene set \nscore"
)
```
